Using SIMD and SIMT vectorization to evaluate sparse chemical kinetic Jacobian matrices and thermochemical source terms

### 2018-06-13 paper.tex
edits through section 3.5.2

\usepackage{todonotes}

@@

\Cref{T:error} reports the maximum $E_{\mathcal{C}}$ and $E_{\mathcal{L}}$ values over all test platforms and vectorization patterns (see~\cref{t:platforms}), sparse and dense (see~\cref{S:sparsity}), as well as \conp/ and \conv/ formulations.
The most stringent filtered error norm ($\mathcal{C} = 10^{20}$) ranges from \numrange[retain-zero-exponent=true]{e-7}{e0}; the largest error is for the \ce{H2}\slash\ce{CO} model.
For this model, $\mathcal{T}$  is smaller than the tolerance of \num{e20}, and hence the error norm considers Jacobian entries smaller than $\mathcal{O}(1)$.
It is noted that for GRI-Mech 3.0, which has a $\mathcal{T}$ roughly an order of magnitude larger, the stringent error norm is significantly smaller: $\mathcal{O}(\num{e-7})$.
Given the intricacy of floating point error evaluation, the use of different languages and OpenCL platforms (the effect of these differences will be explored in \cref{A:per_platform}), and for the general complexity of \texttt{pyJac} it would be exceedingly difficult to pinpoint an exact cause for this phenomenon, as was done in~\cref{s:validation}.
To ensure there is no bug or error present in the Jacobians generated by \texttt{pyJac}, a relaxed filtered error norm ($\mathcal{C} = 10^{15}$) is also presented for each model in~\cref{T:error}.
This relaxed norm is smaller by \numrange{2}{5} orders of magnitude for all models---except GRI-Mech 3.0, where the stringent case has very small error as previously discussed---indicating that accuracy is higher when the smaller Jacobian entries are excluded; this suggests floating point error accumulation is the primary driver of the stringent filtered error norm.

The relative LAPACK error norm---ranging from \textasciitilde\numrange{e-10}{e-14}---re-enforces this finding, as it indicates roughly \numrange{10}{14} digits of accuracy~\cite{Anderson:1999aa}.
The \ce{iC5H11OH} model has the largest LAPACK error norm, likely due to the presence of P-Log\slash Chebyshev reactions and associated complicated derivatives with many logarithms, exponetiations and summations.
Further, the LAPACK error norm does not correlate well with the stringent filtered error norm, e.g., USC-Mech II has the smallest LAPACK error norm (\num{1.137e-14}), but the second largest stringent filtered error norm (\num{1.119e-2}).
Conversely, the model with the largest LAPACK error norm, \ce{iC5H11OH} (with $E_{\mathcal{L}} = \num{1.227e-10}$), has the second smallest stringent filtered error norm (\num{1.363e-3}) , again suggesting that the stringent error norm is influenced by floating point error accumulation.
These findings, along with the individual unit-testing of all chemical source-terms and Jacobian sub-components described in~\cref{s:unittest}, gives high confidence in the correctness of this new version of \texttt{pyJac}.

@@

In general, the Jacobian matrices generated by \texttt{pyJac} are largely sparse with non-zero entries corresponding mainly to species that participate in the same reaction (or non-default efficiency third-body species in a reaction), and mostly dense rows\slash columns corresponding to the temperature and thermodynamic state parameter.
However, the explicit-mass conservation formulation of \texttt{pyJac} can introduce additional non-zero entries in two ways.
First, if the last-species in the model (i.e., the bath-gas) participates directly in a reaction, the derivative of the forward or reverse rate of progress with respect to any other species is non-zero (see~\cref{deriv-s:dri_dnj} of the supplemental material).
Similarly, if the last-species has a third-body efficiency not equal to the default (unity), this will again force all the derivative of the pressure-modification term with respect to other species to be non-zero (see~\cref{deriv-s:dci_thd} of the same).
Either case will result in an undesirable fully dense Jacobian row for all species that have a non-zero net stoichiometic coefficient in such a reaction.

However, the current \texttt{pyJac} allows the user to simply ignore these derivatives (via a command-line switch) and avoid any adverse effects on the Jacobian sparsity.
The rationale behind this choice is that many common implicit integration techniques (e.g., CVODE~\cite{Hindmarsh:2005}) used in solution of chemical kinetic initial value problems are formulated around the assumption that the supplied Jacobian is approximate; this allows the Jacobian and its LU factorization to be reused for multiple internal integration time-steps, accelerating the solution process.
For such solvers, there is no need for the exact form of the Jacobian and the so-called ``approximate'' form is preferable.
\add{Though this might be used as a crude form of preconditioning for such solvers, the primary purpose is merely to increase the resulting sparsity of the Jacobian; the reader is directed to~\cite{MCNENLY2015581} for a more thorough discussion of preconditioners.}
Hence, in this section we will detail the sparsity of both forms of the Jacobian for the chemical models tested.

In~\cref{F:jac_sparsity}, a graphical representation of the sparsity pattern for GRI-Mech 3.0 is displayed.
In particular we note that~\cref{F:jac_sparse_approx} is has several rows that are no longer fully-dense, as result of its approximate form; these rows correspond to species directly interacting with \ce{N2}, largely in GRI-Mech 3.0's nitrogen chemistry reactions.
\cref{T:jac_sparsity} shows the density of the exact and approximate Jacobians for all chemical kinetic models tested in this work.
The smallest model, \ce{H2}\slash\ce{CO} is very dense with \SI{71.4}{\percent} of the exact Jacobian entries non-zero, however this drops to \SI{56.7}{\percent} for GRI-Mech 3.0, and continues to decrease to \SI{28.2}{\percent} for USC-Mech II and just \SI{11.5}{\percent} for the isopentanol model.
The approximate Jacobian assumption drops the density of Jacobian by \textasciitilde\SIrange{3}{7}{\percent} for all models.

\begin{figure}[htb]
  \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{ch4_sparsity_exact.pdf}
      \caption{The ``exact'' Jacobian.}
      \label{F:jac_sparse_exact}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{ch4_sparsity_approximate.pdf}
      \caption{The ``approximate'' Jacobian.}
      \label{F:jac_sparse_approx}
  \end{subfigure}
  \caption{A graphical representation of the sparsity pattern of the chemical kinetic Jacobian generated by \texttt{pyJac} for GRI-Mech 3.0.
	   Black squares indicate a non-zero Jacobian entry, while white square correspond to an empty entry.
	   The numbers indicate the index of the entry in the state vector.}
  \label{F:jac_sparsity}
\end{figure}

Currently, \texttt{pyJac} can utilize two common sparse matrix storage formats~\cite{netlib_templates}, compressed row\slash column storage (CRS\slash CCS), used for ``C'' and ``F''-ordered data respectively.
For convenience we will outline only the CRS format but the CCS format is very similar~\cite{netlib_templates}.
An $N\times N$ CRS matrix is composed of three vectors, a value vector of length $NNZ$---the number of non-zero elements in the Jacobian---which stores the elements of the Jacobian, a row pointer vector---of length $N$---which stores the locations in the value vector that begin a row, and a column index vector---of length $NNZ$---which stores the column indices of the elements in the value vector.

\begin{table}[tbp]
\centering
\begin{tabular}{@{}l S[table-format=2.1] S[table-format=2.1]@{}}
\toprule
Model                 & \multicolumn{1}{c}{Exact Jacobian Density} & \multicolumn{1}{c}{Approximate Jacobian Density} \\
\midrule
\ce{H2}\slash \ce{CO} & \SI{71.4}{\percent} & \SI{68.4}{\percent} \\
GRI-Mech 3.0          & \SI{56.7}{\percent} & \SI{49.8}{\percent} \\
USC-Mech II           & \SI{28.2}{\percent} & \SI{26.4}{\percent} \\
\ce{iC5H11OH}         & \SI{11.5}{\percent} & \SI{7.98}{\percent} \\
\bottomrule
\end{tabular}
\caption{The density of the exact and approximate Jacobians generated by \texttt{pyJac} for the various models studied.}
\label{T:jac_sparsity}
\end{table}

The Jacobian access pattern used by \texttt{pyJac} is fairly irregular; for simplicity we will only discuss looping-structure of species derivatives calculations as these form the bulk of the computation, and have the most challenging Jacobian access patterns.
In general, an outer loop iterates over all reactions of a certain type (e.g., falloff reactions) and calculates the relevant Jacobian subproducts---independent of any particular species---for the reaction (e.g., the derivative of the falloff pressure modification term).
Two inner loops then iterate over the species in a reaction, updating the Jacobian entries for these species as appropriate.
This pattern leads to fairly easily vectorizable code, and efficient Jacobian evaluation as the bulk of the computation is dependent only on the reaction in question, as discussed in our previous work~\cite{Niemeyer:2016aa}.
Generally, this means that a lookup operation is required to find the sparse Jacobian index for any pair of state variables; in some cases this can be avoided, e.g., the rows corresponding to derivatives of the temperature and thermodynamic state parameter source-terms are fully dense in \texttt{pyJac}, and hence no lookup is necessary.
This lookup operation is currently implemented as a simple for-loop, e.g., for a sparse lookup of a pair of indices $(i, j)$ in a CRS matrix, the lookup function searches the column index vector between the values $\texttt{row\_pointer}[i] \ldots \texttt{row\_pointer}[i + 1]$ for $j$, and returns the offset from $\texttt{row\_pointer}[i]$ (or $\num{-1}$ if not found).
As will be explored in~\cref{S:jacobian_results}, this slows down sparse Jacobian evaluation, and might be improved---at the cost of increased constant data usage, a limitation for OpenCL---by a static mapping of the full Jacobian indices to the sparse index (or some null value if the entry is empty).
Additionally, this might be an excellent usage of OpenCL's Image memory type (similar to texture memory in CUDA terminology) and both of these sparse indexing techniques merit future investigation.

\subsection{Performance}
\label{S:results}
The performance studies in this work were run on the platforms listed in~\cref{t:cpus,t:gpus}.
Run times in each case were averaged over ten runs, each using the same set of PaSR conditions utilized in validation.
The OpenMP Jacobian\slash source-term kernels, as well as the OpenMP\slash OpenCL wrapping code---responsible for initializing\slash transferring memory, reading input, etc.---was compiled with \texttt{gcc v5.4.0} on the \avx/\slash\gpunew/ platforms, and \texttt{gcc v4.8.5} on the \sse/\slash\gpuold/ machines.
The optimization level ``\texttt{-O3 -mtune=native}'' was used, and no ``fast-math'' OpenCL optimizations were enabled.
Additionally, the exact form of the Jacobian (as opposed the the ``approximate'' form discussed in~\cref{S:sparsity}) was used in all cases.
Finally, unless stated otherwise: the performance results utilized a single CPU core, the \conp/ assumption, a vector-width of \num{8}\slash\num{128} and ``C''\slash ``F''-ordered data for the CPU\slash GPU cases, respectively; the run times reported are for the number of conditions specified in~\cref{T:error} and include data transfer overhead to\slash from internal buffers used in \texttt{pyJac}.
The effects of choice of vector-width, data-ordering and differences between \conp/ and \conv/ evaluations on the CPU\slash GPU will be explored in~\cref{S:source_results}, while parallel scaling for multiple CPU cores will be exampled in~\cref{S:source_results,S:jacobian_results}.

@@

\Cref{F:cpu_source} explores the performance of the source-term evaluations generated by \texttt{pyJac} on the CPU test platforms listed in~\cref{t:cpus}.
Source-term evaluations---critical in their own right for direct numerical simulations of reactive-flows~\cite{Spafford:2010aa}, among other applications---also give a convenient platform to detail the effects of various code configuration options before investigating the more involved Jacobian evaluation performance.

In~\cref{F:intel_source_nonorm}, the mean run time per initial condition is shown for both the~\avx/ and~\sse/ CPUs, using both Intel OpenCL and OpenMP.
This normalization of the run time by the number of initial conditions tested is chosen to account for the varying numbers of conditions in the PaSR databases for each model (\cref{t:models}).
For both CPUs, the OpenMP implementation is the slowest over all models; interestingly the unvectorized (strictly parallel) Intel OpenCL code is slightly faster than OpenMP in all cases.
As expected, the \avx/ machine is faster than the \sse/ CPU for the strictly parallel cases, \SIrange{1.82}{2.13}{$\times$} for the unvectorized OpenCL case, and \SIrange{1.72}{1.85}{$\times$} for OpenMP.
Additionally, the shallow-vectorized OpenCL code is significantly faster than either the OpenMP or unvectorized OpenCL codes on both processors.

@@

In~\cref{F:intel_source}, the extent of this speedup is detailed; we note that the speedup displayed is calculated per-machine, e.g., the \avx/ shallow-vectorized code speedup is relative to OpenMP on the same CPU such that the speedup of the shallow-vectorized OpenCL code can be readily assessed.
On both machines, the unvectorized OpenCL code is faster than the baseline parallel OpenMP code; by \SIrange{1.30}{1.51}{$\times$} on the \avx/ CPU, and \SIrange{1.25}{1.27}{$\times$} on the \sse/ machine.
Additionally, the shallow-vectorized OpenCL code is \SIrange{2.53}{2.92}{$\times$} and \SIrange{3.40}{4.08}{$\times$} faster than the OpenMP code for the \sse/ and \avx/ machines, respectively.

In contrast,~\cref{F:pocl_source} shows the mean run time per condition of deep, shallow and unvectorized OpenCL codes on the POCL runtime, as compared to OpenMP parallelization.
No speedup is achieved for either vectorization type on either CPU---\add{though we stress that Intel OpenCL runtime achieved vectorization using the same code and hardware}---indeed, the OpenMP case is faster on both CPUs.
The reasons for this lack of vectorization are quite technical,~\delete{and somewhat outside the scope of this work}however \add{personal communications~\cite{pocl_communication} with the developers of POCL revealed that in order to achieve vectorization, changes are required to both the way POCL prepares LLVM \cite{Lattner:2004:LCF:977395.977673} IR-code---specifically, to specialize boundary cases to ensure LLVM recognizes uniform vectorization loop-bounds, even if said bounds \textit{are} uniform in practice---as well as improvements to LLVM's loop-vectorizer: e.g., for proofs of a vector instruction's ability to handle all edge cases identically to the corresponding scalar instruction, handing of branches and conditionals (potentially within POCL instructions themselves), and handling of memory access\slash vector-element extraction patterns.}
As will be discussed in~\cref{S:future}, \add{it is hoped that use of explicit vector-types (to lessen demands on the LLVM-vectorization module) in combination with some of these changes might solve this issue}, but for the moment POCL is still quite useful as a validation tool.

\Cref{F:gpu_source} shows the performance of the source-term evaluation on the GPUs listed in~\cref{t:gpus}.
In~\cref{F:gpu_source_scaling}, the number of initial conditions over which the source-terms are evaluated on the \gpunew/ GPU is varied in order to determine the effect on the mean run time per condition; the run time continues to decrease until \textasciitilde\num{e4} conditions for all chemical models (except the \ce{H2}\slash\ce{CO} model, which is closer to \textasciitilde\num{3e4} conditions), at which point the GPU becomes saturated and performance levels off.
\Cref{F:gpu_source_speedup} shows the speedup in source-term evaluations that the \gpunew/ GPU achieves over the \gpuold/ GPU for the maximum number of conditions in~\cref{F:gpu_source_scaling} with two different vector-widths (a.k.a., the block size in Nvidia terminology), \num{64} and \num{128}; the fastest \gpunew/ case with a vector-width of 128 is \SIrange{1.40}{1.88}{$\times$} faster than the slowest case (\gpuold/ with a vector-width of \num{64}) depending on the chemical model in question.
Additionally,~\cref{F:gpu_source_speedup} shows that varying the vector-width has minimal effect on performance for most of the \gpunew/ and \gpuold/ cases; the GRI-Mech 3.0 and USC-Mech II models show the largest improvements with a vector-width of \num{128}, \textasciitilde\SIrange{10}{18}{\percent} for both GPUs.
This is likely caused by higher occupancy on the GPUs, however it is unclear exactly how Nvidia's OpenCL runtime balances the registers\slash warps per streaming-multiprocessor, as controls occupancy in CUDA~\cite{occupancy}; use of even larger vector-widths (i.e., \num{256}\texttt{+}) on the GPU is a topic that merits future investigation.
\begin{figure}[htbp]
   \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{gpu_source_scaling.pdf}
      \caption{The mean run time per condition for each chemical model on the \gpunew/ GPU as a function of the number of initial conditions tested.}
      \label{F:gpu_source_scaling}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{gpu_source_speedup.pdf}
      \caption{The speedup achieved on the \gpunew/ versus the \gpuold/ GPU; the names correspond to the identifiers listed in~\cref{t:gpus}}
      \label{F:gpu_source_speedup}
  \end{subfigure}
  \caption{\texttt{pyJac} source-term evaluation performance on the Nvidia GPUs}
  \label{F:gpu_source}
\end{figure}

\Cref{f:source_permutate} shows the effect of changing data-ordering patterns, the \conp/ or \conv/ formulation, and the CPU vector-width size on the performance of source-term evaluation in \texttt{pyJac}.
In~\cref{F:source_conpvsconv}, it is seen that choice of the \conp/ or \conv/ formulation has little to no effect on run time for OpenMP as well as the shallow-vectorized\slash unvectorized Intel OpenCL codes on the \avx/ machine.
Generally speaking, the difference between the \conp/ and \conv/ formulations has only a marginal effect on performance regardless of CPU\slash GPU choice.

>>>>

\usepackage[textsize=footnotesize,textwidth=1.5\linewidth]{todonotes}

@@

\Cref{T:error} reports the maximum $E_{\mathcal{C}}$ and $E_{\mathcal{L}}$ values over all test platforms and vectorization patterns (see~\cref{t:platforms}), sparse and dense (see~\cref{S:sparsity}), as well as \conp/ and \conv/ formulations.
The most stringent filtered error norm ($\mathcal{C} = 10^{20}$) ranges from \numrange[retain-zero-exponent=true]{e-7}{e0}; the largest error is for the \ce{H2}\slash\ce{CO} model.
For this model, $\mathcal{T}$ is smaller than the tolerance of \num{e20}, and hence the error norm considers Jacobian entries smaller than $\mathcal{O}(1)$.
GRI-Mech 3.0 has a $\mathcal{T}$ roughly an order of magnitude larger and so the stringent error norm is significantly smaller: $\mathcal{O}(\num{e-7})$.
Given the intricacy of floating-point error evaluation, the use of different languages and OpenCL platforms (the effect of these differences will be explored in \cref{A:per_platform}), and the general complexity of \texttt{pyJac} it would be exceedingly difficult to pinpoint an exact cause for this phenomenon, as was done in~\cref{s:validation}.
To ensure no bugs or errors present in the Jacobians generated by \texttt{pyJac}, relaxed filtered error norms ($\mathcal{C} = 10^{15}$) are also presented for each model in~\cref{T:error}.
This relaxed norm is smaller by \numrange{2}{5} orders of magnitude for all models---except GRI-Mech 3.0, where the stringent case already has small error as previously discussed---which indicates that accuracy is higher when the smaller Jacobian entries are excluded.
\revise{This result suggests that floating-point error accumulation controls the stringent filtered error norm.}

The relative LAPACK error norm---ranging from \textasciitilde\numrange{e-10}{e-14}---re-enforces this finding, as it indicates roughly \numrange{10}{14} digits of accuracy~\cite{Anderson:1999aa}.
The \ce{iC5H11OH} model has the largest LAPACK error norm, likely due to the presence of P-Log\slash Chebyshev reactions and the resulting complicated derivatives with many logarithms, exponetiations, and summations.
Further, the LAPACK error norm does not correlate well with the stringent filtered error norm, e.g., USC-Mech II has the smallest LAPACK error norm (\num{1.137e-14}) but the second-largest stringent filtered error norm (\num{1.119e-2}).
\revise{Conversely, the model with the largest LAPACK error norm, \ce{iC5H11OH} has the second smallest stringent filtered error norm: $E_{\mathcal{L}} = \num{1.227e-10}$ and $E_{\mathcal{C} = 10^{20}} = \num{1.363e-3}$, again suggesting that floating-point error accumulation influences the stringent error norm.}
These findings, along with the individual unit-testing of all chemical source-terms and Jacobian sub-components described in~\cref{s:unittest}, gives high confidence in the correctness of \revise{\texttt{pyJac} v2}.

@@

In general, the Jacobian matrices generated by \texttt{pyJac} are largely sparse with non-zero entries corresponding to species that participate in the same reaction or non-default efficiency third-body species in a reaction, with dense rows\slash columns corresponding to temperature and the thermodynamic state parameter.
However, the explicit-mass conservation formulation of \texttt{pyJac} can introduce additional non-zero entries in two ways.
First, if the last species in the model (i.e., the bath gas) participates directly in any reaction, the derivative of its forward or reverse rate of progress \revise{is non-zero with respect to other species participating in that reaction.}\todo{Nick: check this}{}
Similarly, if the last species has a third-body efficiency not equal to the default (one), \revise{this will again create nonzero derivatives for the pressure-modification term with respect to other species (see the supplemental material).}
Either case will result in a fully dense Jacobian row for all species with a non-zero net stoichiometric coefficient in such a reaction.

However, \texttt{pyJac} v2 allows the user to ignore these derivatives (via a command-line switch) and avoid the adverse effects on Jacobian sparsity.\footnote{Alternatively, one may choose the last species as one that does not participate in any reactions.}
The rationale behind this choice is that many common implicit integration techniques (e.g., CVODE~\cite{Hindmarsh:2005}) used to solve chemical-kinetic initial-value problems are formulated around the assumption that the supplied Jacobian is approximate; this allows the Jacobian and its LU factorization to be reused for multiple internal integration time steps, accelerating the solution process.
\revise{Such solvers do not need the} exact form of the Jacobian and thus the so-called ``approximate'' form is preferable.
\add{Though this might be used as a crude form of preconditioning for such solvers, the primary purpose is merely to increase Jacobian sparsity; McNenly et al.~\cite{MCNENLY2015581} more thorough investigated preconditioners.}
Hence, in this section we will detail the sparsity of both forms of the Jacobian for the chemical models tested.

\Cref{F:jac_sparsity} graphically represents the Jacobian sparsity of GRI-Mech 3.0.
In particular we note that~\cref{F:jac_sparse_approx} has several rows that are no longer fully dense, as result of its approximate form; these rows correspond to species directly interacting with \ce{N2}, largely in GRI-Mech 3.0's nitrogen chemistry reactions.
\cref{T:jac_sparsity} shows the density of the exact and approximate Jacobians for all chemical kinetic models tested in this work.
The smallest model, \ce{H2}\slash\ce{CO}, is very dense with \SI{71.4}{\percent} of the exact Jacobian entries non-zero; this drops to \SI{56.7}{\percent} for GRI-Mech 3.0, continues to decrease to \SI{28.2}{\percent} for USC-Mech II, and is just \SI{11.5}{\percent} for the isopentanol model.
The approximate Jacobian assumption drops the density of Jacobian by \textasciitilde\SIrange{3}{7}{\percent} for all models.

\begin{figure}[htb]
  \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{ch4_sparsity_exact.pdf}
      \caption{The ``exact'' Jacobian.}
      \label{F:jac_sparse_exact}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{ch4_sparsity_approximate.pdf}
      \caption{The ``approximate'' Jacobian.}
      \label{F:jac_sparse_approx}
  \end{subfigure}
  \caption{A graphical representation of the sparsity pattern of the chemical kinetic Jacobian generated by \texttt{pyJac} for GRI-Mech 3.0.
	   Black squares indicate a non-zero Jacobian entry, while white square correspond to an empty entry.
	   The numbers indicate the index of the entry in the state vector.}
  \label{F:jac_sparsity}
  \todo[inline]{the figure on the left is noticeably blurry compared to the other... can we fix that?}
\end{figure}

Currently, \texttt{pyJac} can use two common sparse-matrix storage formats~\cite{netlib_templates}: compressed row storage (CRS) and compressed column storage (CCS), used for ``C'' and ``F''-ordered data respectively.
For brevity we will outline only the CRS format but the CCS format is similar~\cite{netlib_templates}.
An $N\times N$ CRS matrix is stored using three vectors: a value vector of length $N_{NZ}$ (the number of non-zero elements in the Jacobian) that stores the elements of the Jacobian, a row pointer vector of length $N$ that stores the locations in the value vector that begin a row, and a column index vector length $N_{NZ}$ that stores the column indices of the elements in the value vector.

\begin{table}[tbp]
\centering
\begin{tabular}{@{}l S[table-format=2.1] S[table-format=2.1]@{}}
\toprule
Model                 & {Exact Jacobian Density} & {Approximate Jacobian Density} \\
\midrule
\ce{H2}\slash \ce{CO} & \SI{71.4}{\percent} & \SI{68.4}{\percent} \\
GRI-Mech 3.0          & \SI{56.7}{\percent} & \SI{49.8}{\percent} \\
USC-Mech II           & \SI{28.2}{\percent} & \SI{26.4}{\percent} \\
\ce{iC5H11OH}         & \SI{11.5}{\percent} & \SI{7.98}{\percent} \\
\bottomrule
\end{tabular}
\caption{The density of the exact and approximate Jacobians generated by \texttt{pyJac} for the various models studied.}
\label{T:jac_sparsity}
\end{table}

The Jacobian access pattern used by \texttt{pyJac} is fairly irregular; for simplicity we will only discuss looping-structure of species derivatives calculations since these form the bulk of the computation and have the most challenging Jacobian access patterns.
In general, an outer loop iterates over all reactions of a certain type (e.g., falloff reactions) and calculates the relevant Jacobian subproducts---independent of any particular species---for the reaction (e.g., the derivative of the falloff pressure modification term).
Two inner loops then iterate over the species in a reaction, updating the Jacobian entries for these species as appropriate.
This pattern leads to fairly easily vectorizable code and efficient Jacobian evaluation, since the bulk of the computation depends only on the reaction in question, as discussed in our previous work~\cite{Niemeyer:2016aa}.
Generally, this means that a lookup operation is required to find the sparse Jacobian index for any pair of state variables; in some cases this can be avoided, e.g., the rows corresponding to derivatives of the temperature and thermodynamic state parameter source-terms are fully dense in \texttt{pyJac}, and hence no lookup is necessary.
This lookup operation is currently implemented as a simple ``for'' loop, e.g., for a sparse lookup of a pair of indices $(i, j)$ in a CRS matrix, the lookup function searches the column index vector between the values $\texttt{row\_pointer}[i], \ldots, \texttt{row\_pointer}[i + 1]$ for $j$, and returns the offset from $\texttt{row\_pointer}[i]$ (or $\num{-1}$ if not found).
As will be explored in~\cref{S:jacobian_results}, this slows down sparse Jacobian evaluation, and might be improved by a static mapping of the full Jacobian indices to the sparse index (or some null value if the entry is empty).
\revise{However, this would require increased constant-data usage, a limitation for OpenCL.}
Additionally, this might be an excellent usage of OpenCL's Image memory type (similar to texture memory in CUDA terminology).
Both of these sparse indexing techniques merit future investigation.

\subsection{Performance}
\label{S:results}
The performance studies in this work were run on the platforms listed in~\cref{t:cpus,t:gpus}.
Run times in each case were averaged over ten runs, each using the same set of PaSR conditions used in validation.
The OpenMP Jacobian\slash source-term kernels, as well as the OpenMP\slash OpenCL wrapping code (responsible for initializing\slash transferring memory, reading input, etc.) was compiled with \texttt{gcc v5.4.0} on the \avx/\slash\gpunew/ platforms and \texttt{gcc v4.8.5} on the \sse/\slash\gpuold/ machines.
The optimization level ``\texttt{-O3 -mtune=native}'' was used and no ``fast math'' OpenCL optimizations were enabled.
Additionally, the exact form of the Jacobian (as opposed the the ``approximate'' form discussed in~\cref{S:sparsity}) was used in all cases.
Finally, unless stated otherwise: the performance results used a single CPU core, the \conp/ assumption, a vector width of \num{8}\slash\num{128}, and ``C''\slash ``F''-ordered data for the CPU\slash GPU cases, respectively; the run times reported are for the number of conditions specified in~\cref{T:error} and include data-transfer overhead to\slash from internal buffers used in \texttt{pyJac}.
The effects of choice of vector width, data ordering, and differences between \conp/ and \conv/ evaluations on the CPU\slash GPU will be explored in~\cref{S:source_results}, while parallel scaling for multiple CPU cores will be exampled in~\cref{S:source_results,S:jacobian_results}.\todo{is ``exampled'' really the word you want here?}{}

@@

\Cref{F:cpu_source} explores the performance of the source-term evaluations generated by \texttt{pyJac} on the CPU test platforms listed in~\cref{t:cpus}.
Source-term evaluations---critical in their own right for direct numerical simulations of reactive-flows~\cite{Spafford:2010aa}, among other applications---also provide a convenient platform to detail the effects of various code configuration options before investigating the more involved Jacobian evaluation performance.

\Cref{F:intel_source_nonorm} shows the mean run time per initial condition for both the~\avx/ and~\sse/ CPUs, using Intel OpenCL and OpenMP.
This normalization of the run time by the number of initial conditions tested is chosen to account for the varying numbers of conditions in the PaSR databases for each model (\cref{t:models}).
For both CPUs, the OpenMP implementation is the slowest for all models; interestingly, the unvectorized (i.e., strictly parallel) Intel OpenCL code is slightly faster than OpenMP in all cases.
As expected, the \avx/ machine is faster than the \sse/ CPU for the strictly parallel cases, performing \SIrange{1.82}{2.13}{$\times$} and \SIrange{1.72}{1.85}{$\times$} faster for the unvectorized OpenCL case and OpenMP, respectively.
Additionally, the shallow-vectorized OpenCL code performs significantly faster than either the OpenMP or unvectorized OpenCL codes on both processors.

@@

\Cref{F:intel_source} details the extent of this speedup; the speedup displayed is calculated per-machine, e.g., the \avx/ shallow-vectorized code speedup is relative to OpenMP on the same CPU\revisedelete{ such that the speedup of the shallow-vectorized OpenCL code can be readily assessed}.
On both machines, the unvectorized OpenCL code is faster than the baseline parallel OpenMP code, by \SIrange{1.30}{1.51}{$\times$} on the \avx/ CPU and \SIrange{1.25}{1.27}{$\times$} on the \sse/ machine.
Additionally, the shallow-vectorized OpenCL code is \SIrange{2.53}{2.92}{$\times$} and \SIrange{3.40}{4.08}{$\times$} faster than the OpenMP code for the \sse/ and \avx/ machines, respectively.

In contrast,~\cref{F:pocl_source} shows the mean run time per condition of deep, shallow, and unvectorized OpenCL codes using the POCL runtime, as compared with OpenMP parallelization.
No speedup is achieved for either vectorization type on either CPU---indeed, the OpenMP case is faster on both CPUs\add{, though we stress that Intel OpenCL runtime achieves vectorization using the same code and hardware}.
The reasons for this lack of vectorization with POCL are quite technical~\delete{ and somewhat outside the scope of this work}; however, \add{personal communication with the developers of POCL revealed that to achieve vectorization, changes are required to both the way POCL prepares LLVM \cite{Lattner:2004:LCF:977395.977673} IR-code---specifically, to specialize boundary cases to ensure LLVM recognizes uniform vectorization loop bounds, even if said bounds \textit{are} uniform in practice---as well as improvements to LLVM's loop-vectorizer: e.g., for proofs of a vector instruction's ability to handle all edge cases identically to the corresponding scalar instruction, handing of branches and conditionals (potentially within POCL instructions themselves), and handling of memory access\slash vector-element extraction patterns~\cite{pocl_communication}.}\todo{this is one super long runon sentence---needs to be broken up and possibly simplified.}{}
As will be discussed in~\cref{S:future}, \add{we hope that using explicit vector types (to lessen demands on the LLVM-vectorization module) in combination with some of these changes might solve this issue}, but for the moment POCL is still quite useful as a validation tool.

\Cref{F:gpu_source} shows the performance of evaluating source terms on the GPUs listed in~\cref{t:gpus}.
\revise{\Cref{F:gpu_source_scaling} investigates how the number of initial conditions evaluated affects the mean run time per condition on the \gpunew/ GPU};
the run time decreases until around \textasciitilde\num{e4} conditions for all chemical models, at which point the GPU becomes saturated and performance levels off.
The performance plateaus slightly later for the \ce{H2}\slash\ce{CO} model compared with the others.
\Cref{F:gpu_source_speedup} shows the speedup in source-term evaluation that the \gpunew/ GPU achieves over the \gpuold/ GPU for the maximum number of conditions in~\cref{F:gpu_source_scaling} with two vector widths (i.e., GPU block size), \num{64} and \num{128}.
The best \gpunew/ case with a vector width of 128 is \SIrange{1.40}{1.88}{$\times$} faster than the slowest case (\gpuold/ with a vector-width of 64) depending on the chemical model in question.
\Cref{F:gpu_source_speedup} also shows that varying the vector-width minimally affects performance for most of the \gpunew/ and \gpuold/ cases; the GRI-Mech 3.0 and USC-Mech II models show the largest improvements with a vector width of 128: \textasciitilde\SIrange{10}{18}{\percent} for both GPUs.
This is likely caused by higher occupancy on the GPUs, but it is unclear exactly how Nvidia's OpenCL runtime balances the registers\slash warps per streaming-multiprocessor, as controls occupancy in CUDA~\cite{occupancy};\todo{this last fragment is missing something}{}
use of even larger vector widths (i.e., $\geq$\num{256}) on the GPU is a topic that merits future investigation.\todo[inline]{I feel like we should pull this last bit, since someone could ask why you didn't just run with 256}

\begin{figure}[htbp]
   \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{gpu_source_scaling.pdf}
      \caption{The mean run time per condition for each chemical model on the \gpunew/ GPU as a function of the number of initial conditions tested.}
      \label{F:gpu_source_scaling}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{gpu_source_speedup.pdf}
      \caption{The speedup achieved on the \gpunew/ versus the \gpuold/ GPU; the names correspond to the identifiers listed in~\cref{t:gpus}}
      \label{F:gpu_source_speedup}
  \end{subfigure}
  \caption{\texttt{pyJac} source-term evaluation performance on the Nvidia GPUs}
  \label{F:gpu_source}
\end{figure}

\Cref{f:source_permutate} shows how changing data-ordering patterns, the \conp/ or \conv/ formulation, and the CPU vector width affect the performance of source-term evaluation in \texttt{pyJac}.
Per~\cref{F:source_conpvsconv}, we see that the choice of \conp/ or \conv/ formulation has little to no effect on run time for OpenMP as well as the shallow-vectorized\slash unvectorized Intel OpenCL codes on the \avx/ machine.
Generally speaking, the difference between the \conp/ and \conv/ formulations only marginally affects performance regardless of CPU\slash GPU choice.
